---
title: "Exploratory data analysis of a whole blood microarray data set of healthy individuals in the SHIP-TREND consortium"
author: "Eddie Cano-Gamez"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This markdown contains the pre-processing and exploratory data analysis of a publicly available microarray data set of whole blood from 991 healthy individuals from the Pomerania region in Germany. These individuals were recruited as a part of the SHIP-TREND consortium.

RNA was extracted from whole blood using the PAXgene Blood RNA System, and gene epxression measurement was performed with an Illumina HumanHT-12 v3 microarray. For more details on the generation and processing of this data set, refer to the SHIP-TREND cohort website or Westra et al.; Nature Genetics:

https://www.nature.com/articles/ng.2756

## Loading libraries
```{r load_libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(SummarizedExperiment)
library(illuminaHumanv3.db)
```

## Loading data
Microarray data:
```{r load_data, message=FALSE, warning=FALSE}
ship_trend_gex <- read.table("../data/SHIP-TREND/GSE36382_non_normalized.txt.gz", 
                             sep="\t", header=T, row.names = 1)

```

Illumina HumanHT probe mappings:
```{r load_IlluminaHumanHT_probe_mappings, message=FALSE, warning=FALSE}
probe_annotations <- read_tsv("../data/SHIP-TREND/GPL6947_HumanHT-12_V3_0_R1_11283641_A.bgx.gz", skip=8)
```

## Filtering probes by detection reliability
The histogram below shows the number of microarray probes which were reliably detected for every given number of individuals. Here, reliable detection is defined as the probe reaching a detection P value lower or equal than 0.01.
```{r plot_probes_by_number_of_individual_passing_pval_filter, message=FALSE, warning=FALSE, echo=FALSE}
hist(rowSums(ship_trend_gex[,grepl("DetectionPvalue",colnames(ship_trend_gex))] < 0.01), 
     xlab="Number of samples where probe was reliably detected (p value < 0.01)", 
     ylab="Number of probes", 
     main="Microarray probes QC")
```

Based on the distribution shown above, only probes which were reliably detected in at least 200 samples are kept for further analyses.
```{r remove_unreliable_probes, message=FALSE, warning=FALSE}
high_quality_probes <- rownames(ship_trend_gex)[rowSums(ship_trend_gex[,grepl("DetectionPvalue",colnames(ship_trend_gex))] < 0.01) > 200]

ship_trend_gex <- ship_trend_gex[high_quality_probes, 
                                 grep("DetectionPvalue",colnames(ship_trend_gex), invert = T)]
```

## Log transforming
We now apply a log2 transformation so as to bring the intensity measurements to a more Gaussian distribution.

```{r log_transform, message=FALSE, warning=FALSE}
ship_trend_gex <- log2(ship_trend_gex)
```

## Collapsing probe measurements by gene
Let's use the probe information file from Illumina to retrieve annotations for all probes that passed QC filters.
```{r find_probes_in_dataset, message=FALSE, warning=FALSE}
probe_annotations <- data.frame(filter(probe_annotations, Probe_Id %in% high_quality_probes))
```

Based on these annotatinos, we then identify all genes present in the study
```{r list_genes_in_study, message=FALSE, warning=FALSE}
genes_in_study <- unique(probe_annotations$ILMN_Gene)
```

We can now collapse probe measurements at the gene level (i.e. average the measurements from all probes mapping to a given gene)
```{r collapse_probes_per_gene, message=FALSE, warning=FALSE}
ship_trend_gex_per_gene <- sapply(genes_in_study, FUN = function(gene){
  probe_ids <- as.character(probe_annotations$Probe_Id[probe_annotations$ILMN_Gene==gene])
  mean_expression <- colMeans(ship_trend_gex[probe_ids,])
  return(mean_expression)
})
ship_trend_gex_per_gene <- data.frame(t(ship_trend_gex_per_gene))
```

## Creating a gene annotation table
Fetching the Ensembl ID for each probe which passed QC based on the gene it maps to.
```{r map_probes_to_ensembl_ids, message=FALSE, warning=FALSE}
probe2ensembl <- as.list(
  illuminaHumanv3ENSEMBL[intersect(probe_annotations$Probe_Id, mappedkeys(illuminaHumanv3ENSEMBL))]
)

ensembl_ids <- c()
for(i in probe_annotations$Probe_Id){
  if(i %in% names(probe2ensembl)){
    gene_id <- probe2ensembl[[i]][[1]]
  } else{
    gene_id <- NA
  }
  ensembl_ids <- c(ensembl_ids, gene_id)
}
probe_annotations$Ensembl_ID <- ensembl_ids
```

Fetching annotations for all genes in the study
```{r fetch_gene_annotations, message=FALSE, warning=FALSE}
gene_annotations <- probe_annotations[!duplicated(probe_annotations$ILMN_Gene), c("ILMN_Gene","Ensembl_ID","Chromosome","Source","RefSeq_ID","Symbol","Protein_Product")]
rownames(gene_annotations) <- gene_annotations$ILMN_Gene
```

## Creating a Summarized Experiment objetct
Creating a SummarizedExperiment containing expression values per gene and gene annotations.
```{r create_summarized_experiment, message=FALSE, warning=FALSE}
SHIP_TREND <- SummarizedExperiment(assays = ship_trend_gex_per_gene,
                                   rowData = gene_annotations)
```

## Dimensionality reduction and visualisation
Inspecting the relationship between expression variance and expression level.
```{r estiamte_gene_variances, message=FALSE, warning=FALSE}
gene_variances <- data.frame(mean_expression = rowMeans(assay(SHIP_TREND)),
                             variance = rowVars(as.matrix(assay(SHIP_TREND))))
```

```{r find_top_variable_genes, message=FALSE, warning=FALSE}
gene_variances <- gene_variances[order(-gene_variances$variance),]
gene_variances$rank <- 1:nrow(gene_variances)
gene_variances$top_variable <- FALSE
gene_variances$top_variable[1:1000] <- TRUE
gene_variances <- gene_variances[rownames(SHIP_TREND),]
```

```{r plot_variance_vs_expression, message=FALSE, warning=FALSE, echo=FALSE}
ggplot(gene_variances, aes(x=mean_expression, y=variance)) + 
  geom_point(aes(color=top_variable), size=0.75) +
  scale_color_manual(values=c("#8da0cb","#fc8d62")) +
  xlab("Mean expression") +
  ylab("Variance in expression") +
  theme_bw() +
  theme(panel.grid = element_blank())
```

Performing principal component analysis (PCA) based on all genes.
```{r perform_pca, message=FALSE, warning=FALSE}
pca_ship_trend <- prcomp(t(assay(SHIP_TREND)))   
```

Fetching PCA coordinates and the variance explained by each component.
```{r fetch_pc_coords, message=FALSE, warning=FALSE}
pc_coords_ship_trend <- data.frame(cbind(pca_ship_trend$x,colData(SHIP_TREND)))
pc_variance_ship_trend <- pca_ship_trend$sdev^2/sum(pca_ship_trend$sdev^2)
```

As shown by the scree plot below, most variance is captured by the first principal component.
```{r plot_variance_explained_per_pc, message=FALSE, warning=FALSE, echo=FALSE}
plot(pc_variance_ship_trend[1:50], 
     type="b", cex=0.5, 
     xlab="Principal Component", 
     ylab="Proportion of variance explained")
```

The samples in this study look as follows in PCA space. There seems to be some degree of heterogeneity, with two clear groups of smaples.
```{r plot_pca, message=FALSE, warning=FALSE, echo=FALSE}
ggplot(pc_coords_ship_trend, aes(x=PC1, y=PC2)) + 
  geom_point(size=3) +
  xlab(paste("PC1 (", round(pc_variance_ship_trend[1]*100,2), "% variance explained)")) +
  ylab(paste("PC2 (", round(pc_variance_ship_trend[2]*100,2), "% variance explained)")) +
  theme_bw() +
  theme(panel.grid = element_blank())
```

## Exporting data set
Finally, the SummarizedExperiment object containing this data is exported as an RDS file.
```{r export_as_rds, eval=FALSE}
saveRDS(SHIP_TREND, "../data/SHIP-TREND/SHIP-TREND_microarray_averaged-per-gene.rds")
```

## R session details
```{r r_session, message=FALSE, warning=FALSE}
sessionInfo()
```
