---
title: "Exploratory data analysis of a whole blood microarray data set of healthy individuals"
author: "Eddie Cano-Gamez"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This markdown contains simple pre-processing and exploratory data analysis of a publicly available microarray data set of whole blood from unrelated individuals of Finnish ancestry. These individuals were part of the FINRISK study. Thus, they were relatively healthy, but with a slightly elevated risk for cardiovascular disease. 

RNA was extracted from whole blood using the PAXgene Blood RNA System, and gene epxression measurement was performed with an Illumina HumanHT-12 microarray (i.e. polyA-based method). For more details on the generation and processing of this data set, refer to Inouye et al, 2010; PLOS Genetics:

https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1001113

## Loading libraries
```{r load_libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(SummarizedExperiment)
library(annotables)
library(illuminaHumanv4.db)
```

## Loading data
Microarray data:
```{r load_data, message=FALSE, warning=FALSE}
inouye_gex <- read.table("../data/Inouye-et-al_2010_PLOS-genetics/E-TABM-1036.processed.txt",
                         sep = "\t",
                         header = T, 
                         row.names = 1)
```

Illumina HumanHT probe mappings:
```{r load_IlluminaHumanHT_probe_mappings, message=FALSE, warning=FALSE}
probe_mappings <- read_tsv("/well/jknight/Sepsis/Gene_Expression/Microarray_combined_816/HumanHT-12_V4_0_R2_15002873_B.txt")
```

## Collapsing probe measurements by gene
Keeping only information for probes present in the data set
```{r find_probes_in_dataset, message=FALSE, warning=FALSE}
probe_mappings <- filter(probe_mappings, Probe_Id %in% rownames(inouye_gex))
probe_mappings <- data.frame(probe_mappings)
```

Adding Ensembl IDs to all probes based on the gene they map to
```{r map_probes_to_ensembl_ids, message=FALSE, warning=FALSE}
probe2ensembl <- as.list(
  illuminaHumanv4ENSEMBL[intersect(probe_mappings$Probe_Id, mappedkeys(illuminaHumanv4ENSEMBL))]
)

ensembl_ids <- c()
for(i in probe_mappings$Probe_Id){
  if(i %in% names(probe2ensembl)){
    gene_id <- probe2ensembl[[i]][[1]]
  } else{
    gene_id <- NA
  }
  ensembl_ids <- c(ensembl_ids, gene_id)
}
probe_mappings$Ensembl_ID <- ensembl_ids
```

Identifying all genes present in the study
```{r list_genes_in_study, message=FALSE, warning=FALSE}
genes_in_study <- unique(probe_mappings$ILMN_Gene)
```

Collapsing probe measurements at the gene level (i.e. averaging measurements from all probes mapping to a given gene)
```{r collapse_probes_per_gene, message=FALSE, warning=FALSE}
expression_per_gene <- sapply(genes_in_study, FUN = function(gene){
  probe_ids <- as.character(probe_mappings$Probe_Id[probe_mappings$ILMN_Gene==gene])
  mean_expression <- colMeans(inouye_gex[probe_ids,])
  return(mean_expression)
})
expression_per_gene <- data.frame(t(expression_per_gene))
```

Fetching Ensembl annotations for all genes in the study
```{r fetch_gene_annotations, message=FALSE, warning=FALSE}
gene_annotations <- probe_mappings[!duplicated(probe_mappings$ILMN_Gene),
                                c("ILMN_Gene","Ensembl_ID","Chromosome","Source","RefSeq_ID","Symbol","Protein_Product")]
rownames(gene_annotations) <- gene_annotations$ILMN_Gene
```

## Creating a Summarized Experiment objetct
Creating a SummarizedExperiment containing expression values per gene and gene annotations.
```{r create_summarized_experiment, message=FALSE, warning=FALSE}
inouye_microarray <- SummarizedExperiment(assays = expression_per_gene,
                                          rowData = gene_annotations)
```

## Dimensionality reduction and visualisation
Inspecting the relationship between expression variance and expression level.
```{r estiamte_gene_variances, message=FALSE, warning=FALSE}
gene_variances <- data.frame(mean_expression = rowMeans(assay(inouye_microarray)),
                             variance = rowVars(as.matrix(assay(inouye_microarray))))
```

```{r find_top_variable_genes, message=FALSE, warning=FALSE}
gene_variances <- gene_variances[order(-gene_variances$variance),]
gene_variances$rank <- 1:nrow(gene_variances)
gene_variances$top_variable <- FALSE
gene_variances$top_variable[1:1000] <- TRUE
gene_variances <- gene_variances[rownames(inouye_microarray),]
```

```{r plot_variance_vs_expression, message=FALSE, warning=FALSE, echo=FALSE}
ggplot(gene_variances, aes(x=mean_expression, y=variance)) + 
  geom_point(aes(color=top_variable), size=0.75) +
  scale_color_manual(values=c("#8da0cb","#fc8d62")) +
  xlab("Mean expression") +
  ylab("Variance in expression") +
  theme_bw() +
  theme(panel.grid = element_blank())
```

Performing principal component analysis (PCA) based on all genes.
```{r perform_pca, message=FALSE, warning=FALSE}
pca_inouye <- prcomp(t(assay(inouye_microarray)))   
```

Fetching PCA coordinates and the variance explained by each component.
```{r fetch_pc_coords, message=FALSE, warning=FALSE}
pc_coords_inouye <- data.frame(cbind(pca_inouye$x,colData(inouye_microarray)))
pc_variance_inouye <- pca_inouye$sdev^2/sum(pca_inouye$sdev^2)
```

Most variance is explained by the first 5 PCs.
```{r plot_variance_explained_per_pc, message=FALSE, warning=FALSE, echo=FALSE}
plot(pc_variance_inouye[1:50], 
     type="b", cex=0.5, 
     xlab="Principal Component", 
     ylab="Proportion of variance explained")
```

The samples in this study look as follows in PCA space. There seems to be some degree of heterogeneity, with potentially two groups of smaples.
```{r plot_pca, message=FALSE, warning=FALSE, echo=FALSE}
ggplot(pc_coords_inouye, aes(x=PC1, y=PC2)) + 
  geom_point(size=3) +
  xlab(paste("PC1 (", round(pc_variance_inouye[1]*100,2), "% variance explained)")) +
  ylab(paste("PC2 (", round(pc_variance_inouye[2]*100,2), "% variance explained)")) +
  theme_bw() +
  theme(panel.grid = element_blank())
```

## Exporting data set
Finally, the SummarizedExperiment object containing this data is exported as an RDS file.
```{r export_as_rds, eval=FALSE}
saveRDS(inouye_microarray, "../data/Inouye-et-al_2010_PLOS-genetics/Inouye-et-al_microarray_averaged-per-gene.rds")
```

## R session details
```{r r_session, message=FALSE, warning=FALSE}
sessionInfo()
```
