---
title: Exploratory data analysis of sepsis whole blood transcriptomes from the MARS cohort
  cohort
author: "Eddie Cano-Gamez"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading libraries
```{r load_libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(SummarizedExperiment)
library(annotables)
```

## Loading data
Microarray data
```{r load_data, message=FALSE, warning=FALSE}
mars_gex <- read.csv("/well/jknight/justin/sepsis/mars/mars_all_expression.csv", header = T, row.names = 1)
mars_metadata <- read.csv("../data/mars_gex/MARS-full-metadata_GEO.csv", header=T, row.names = 1)
```

Affymetrix probe annotations
```{r load_probe_annotations, message=FALSE, warning=FALSE}
probe_annotations <- read.csv("/well/jknight/justin/sepsis/mars/affymetrix_human_gene_u219_array_metadata.csv", header=T, row.names = 1)
```

## Pre-processing data
Finding genes in study
```{r find_genes_in_study, message=FALSE, warning=FALSE}
probe_annotations <- probe_annotations[rownames(mars_gex),]
genes_in_study <- unique(probe_annotations$Symbol)
```

Collapsing probe measurements per gene
```{r collapse_probes_per_gene, message=FALSE, warning=FALSE}
mars_gex_per_gene <- sapply(genes_in_study, FUN=function(gene){
  probe_ids <- rownames(probe_annotations)[probe_annotations$Symbol == gene]
  mean_gene_exp <- colMeans(mars_gex[probe_ids,])
  return(mean_gene_exp)
})

mars_gex_per_gene <- data.frame(t(mars_gex_per_gene))
```

Removing probes mapping to multiple genes
```{r remove_multimapping_probes, message=FALSE, warning=FALSE}
mars_gex_per_gene <- mars_gex_per_gene[!grepl("///",rownames(mars_gex_per_gene)),]
```

Creating gene annotations
```{r create_gene_annotations, message=FALSE, warning=FALSE}
gene_annotations <- probe_annotations[probe_annotations$Symbol %in% rownames(mars_gex_per_gene),c("Symbol","Ensembl.Entrez.Gene","chrom")]
gene_annotations <- gene_annotations[!duplicated(gene_annotations$Symbol),]

colnames(gene_annotations) <- c("gene_name","gene_id","chr")
rownames(gene_annotations) <- gene_annotations$gene_name

gene_annotations$chr <- gsub("chr","",gene_annotations$chr)
gene_annotations <- gene_annotations[rownames(mars_gex_per_gene),]
```

Creating a summarized experiment object with expression values, sample metadata, and gene annotations
```{r create_summarized_experiment, message=FALSE, warning=FALSE}
mars <- SummarizedExperiment(assays = mars_gex_per_gene, 
                             rowData = gene_annotations, 
                             colData = mars_metadata)

mars
```

## Exploratory data analysis
Performing principal component analysis
```{r perform_pca, message=FALSE, warning=FALSE}
pca_mars <- prcomp(t(assay(mars)))

pc_coords_mars <- data.frame(cbind(pca_mars$x),colData(mars))
```


Visualising the distribution of samples belonging to different MARS endotypes in PCA space.
```{r plot_pca, message=FALSE, warning=FALSE, echo=FALSE}
ggplot(pc_coords_mars, aes(x=PC1, y=PC2)) +
  geom_point(aes(color=Endotype_cohort), size=3) +
  theme_classic()

ggplot(pc_coords_mars, aes(x=PC1, y=PC2)) +
  geom_point(aes(color=Endotype_class), size=3) +
  theme_classic()

ggplot(pc_coords_mars, aes(x=PC1, y=PC2)) +
  geom_point(aes(color=Group), size=3) +
  theme_classic()
```

## Exporting results
Exporting summarized experiment object as an R data set.
```{r export_data, eval=FALSE}
saveRDS(mars, "../data/mars_gex/mars_microarray_averaged-by-gene.rds")
```

## R session
```{r r_session, message=FALSE, warning=FALSE}
sessionInfo()
```

